A note on the instability and degeneracy of deep learning models

### 2016-11-17 writing/note.Rmd
try to do nordman's ideas

---
title: "A note on the instability and degeneracy of deep learning models"
author: "Andee Kaplan, Daniel Nordman, and Stephen Vardeman"
date: "October, 2016"
linestretch: 2
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: latex/header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  We introduce *Finitely Supported Finite Sequence*, or FSFS models, a broad class of discrete space models that are defined in relation to how a probability model is built to incorporate more variables. These models are defined to be in discrete finite space with model support equal the sample space and with a sequence of expanding model parameter vectors that potentially grow size as the dimension of the data become larger. In this paper, we present results that define the notion of model instability as well as near-degeneracy within the framework of FSFS models in addition to demonstrating example models of the FSFS class.
---

@@

A probability model exhibits instability if small changes in a data outcome result in large, perhaps unanticipated changes in probability. As data structures become larger and more complex, there is increasing interest in identifying instability in potential model specifications for such data (e.g., models for
random graphs or networks). We consider the problem of  quantifying instability for general probability models defined on sequences of observations, where each sequence of a given length $N$ has a finite number of possible outcomes. As the number of available observations increases, a sequence of probability models results to accommodate the data of expanding dimension.

Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than the length $n$ of the data sequence. In this case, a one component change in the data sequence can provably shift probability
by orders of magnitude. Also, as the measure of instability becomes more extreme, the resulting probability models are shown  to slip into degeneracy, thereby placing all probability
on arbitrarily small portions of the sample space. These results on instability apply to large classes of exponential models commonly used in random graphs, network analysis, and machine learning contexts.

@@

Section \ref{examples} provides several examples of FSFS models commonly used in graph/network analysis and machine learning (i.e., deep learning models) among other application areas. Section \ref{instability-results} establishes formal results regarding the model propriety of FSFS models. These results help to measure, and qualify the effects of, model instability that can arise when certain maximal probabilities in such models are too extreme relevant to the length $N$ of the data sample. **TODO: Somewhere here or in Section 3 we need a brief literature review connecting instablity here to previous work (e.g., Schweinberger) and related notions of degeneracy (e.g., Handcock)**. Lastly, Section \ref{implications} emphasizes the implications of our model propriety results.

@@

By above definition, $X_1, \dots, X_N$ are discrete with a finite sample space. Let $\boldsymbol \theta = (\lambda_1, \dots, \lambda_k, 0, \dots, 0)$, where the first $k$ elements are equal to the $k$ elements of $\boldsymbol \lambda$ and for $i = k + 1, \dots, q(N)$, $\theta_i = 0$. Further, because $p_{\boldsymbol \lambda}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$ by properties of the exponential function, the discrete exponential family models defined in this section are a special case of the FSFS model. **TODO: need to briefly add where such exponential models arise (e.g., spatial data, random graphs, network data) to indicate wide-spread applications**

>>>>

---
title: "A note on the instability and degeneracy of deep learning models"
author: "Andee Kaplan, Daniel Nordman, and Stephen Vardeman"
date: "October, 2016"
linestretch: 2
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: latex/header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A probability model exhibits instability if small changes in a data outcome result in large, perhaps unanticipated changes in probability. As data structures become larger and more complex, there is increasing interest in identifying instability in potential model specifications for such data (e.g., models for random graphs or networks). We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of a given length $N$ has a finite number of possible outcomes. As the number of available observations increases, a sequence of probability models results to accommodate the data of expanding dimension.
  Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than the length $N$ of the data sequence. In this case, a one component change in the data sequence can provably shift probability by orders of magnitude. Also, as the measure of instability becomes more extreme, the resulting probability models are shown  to slip into degeneracy, thereby placing all probability on arbitrarily small portions of the sample space. These results on instability apply to large classes of exponential models commonly used in random graphs, network analysis, and machine learning contexts.
---

@@

A probability model exhibits instability if small changes in a data outcome result in large, perhaps unanticipated changes in probability. As data structures become larger and more complex, there is increasing interest in identifying instability in potential model specifications for such data (e.g., models for random graphs or networks). The concept of instability, introduced in the field statistical physics by @ruelle1999statistical, was extended to include a notion of detection and quantification by @schweinberger2011instability. We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of a given length $N$ has a finite number of possible outcomes. As the number of available observations increases, a sequence of probability models results to accommodate the data of expanding dimension.

Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than the length $N$ of the data sequence. In this case, a one component change in the data sequence can provably shift probability by orders of magnitude. Also, as the measure of instability becomes more extreme, the resulting probability models are shown to slip into degeneracy, thereby placing all probability on arbitrarily small portions of the sample space. This relationship between instability and degeneracy, as defined by @handcock2003assessing, was described by @schweinberger2011instability. We extend these results on instability to large classes of exponential models commonly used in random graphs, network analysis, and machine learning contexts.

@@

Section \ref{examples} provides several examples of FSFS models commonly used in graph/network analysis and machine learning (i.e., deep learning models) among other application areas. Section \ref{instability-results} establishes formal results regarding the model propriety of FSFS models. These results help to measure, and qualify the effects of, model instability that can arise when certain maximal probabilities in such models are too extreme relevant to the length $N$ of the data sample. Lastly, Section \ref{implications} emphasizes the implications of our model propriety results.

@@

By above definition, $X_1, \dots, X_N$ are discrete with a finite sample space. Let $\boldsymbol \theta = (\lambda_1, \dots, \lambda_k, 0, \dots, 0)$, where the first $k$ elements are equal to the $k$ elements of $\boldsymbol \lambda$ and for $i = k + 1, \dots, q(N)$, $\theta_i = 0$. Further, because $p_{\boldsymbol \lambda}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$ by properties of the exponential function, the discrete exponential family models defined in this section are a special case of the FSFS model. Such exponential models arise in spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social, @handcock2003assessing], and even in binomial sampling with $N$ iid Bernoulli random variables.
